### 2024-07-28 mixed-models.qmd
20240728 - fit indices

@@

#### Linear Regression Model {#sec-fantasyPointsByAgeExperienceModelsRegression}

```{r}
pointsPerSeason_regressionModel <- lm(
  fantasy_points ~ 1,
  data = player_stats_seasonal_offense_subset,
  na.action = "na.exclude"
)

summary(pointsPerSeason_regressionModel)
summary(pointsPerSeason_regressionModel)$r.squared
AIC(pointsPerSeason_regressionModel)
MuMIn::AICc(pointsPerSeason_regressionModel)
```

@@

After fitting several models, we now must compare their fit to determine which model fits "best" while also considering parsimony.
Parsimonious models are more likely to be true and more likely to generalize to other samples, because more complex models are more likely to overfit the data.
Thus, more complex models will almost always fit better than simpler models.
Thus, we are not just interested in whether a more complex model fits better than the simpler model; we also care about whether the more complex model fits *significantly* better than the simpler model given its additional complexity.
For evaluating and comparing models, we examine the [likelihood ratio test](#sec-fantasyPointsByAgeExperienceModelsAgeCompareLRT), the [Akaike Information Criterion](#sec-fantasyPointsByAgeExperienceModelsAgeCompareAIC) (AIC), the [corrected AIC](#sec-fantasyPointsByAgeExperienceModelsAgeCompareAICc) (AICc), the [Bayesian Information Criterion](#sec-fantasyPointsByAgeExperienceModelsAgeCompareBIC) (BIC), [$R^2$](#sec-fantasyPointsByAgeExperienceModelsAgeCompareRsquared), and [deviance](#sec-fantasyPointsByAgeExperienceModelsAgeCompareDeviance).

@@

```{r}
lmVsMixedModel <- list(
  "linearRegression" = pointsPerSeason_regressionModel,
  "randomIntercepts" = pointsPerSeason_randomIntercepts
)

lmAndMixedModels <- list(
  "linearRegression" = pointsPerSeason_regressionModel,
  "randomIntercepts" = pointsPerSeason_randomIntercepts,
  "position" = pointsPerSeason_position,
  "fixedLinear" = pointsPerSeason_positionAgeFixedLinearSlopes,
  "randomLinear" = pointsPerSeason_positionAgeRandomLinearSlopes,
  "randomLinearFixedQuadratic" = pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes,
  "randomLinearFixedQuadraticInteraction" = pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction
)

@@

```{r}
AIC(
  pointsPerSeason_regressionModel,
  pointsPerSeason_randomIntercepts,
  pointsPerSeason_positionAgeFixedLinearSlopes,
  pointsPerSeason_positionAgeRandomLinearSlopes,
  pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes,
  pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction
  )

@@

```{r}
BIC(
  pointsPerSeason_regressionModel,
  pointsPerSeason_randomIntercepts,
  pointsPerSeason_positionAgeFixedLinearSlopes,
  pointsPerSeason_positionAgeRandomLinearSlopes,
  pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes,
  pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction
  )

@@

AICcmodavg::bictab(mixedModels) # throws error (can't mix lm with lmer)
#bbmle::AICctab(mixedModels) # throws error (different numbers of observations)

@@

```{r}
summary(pointsPerSeason_regressionModel)$r.squared
MuMIn::r.squaredGLMM(pointsPerSeason_randomIntercepts)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeFixedLinearSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction)
```

@@

```{r}
deviance(pointsPerSeason_regressionModel)
deviance(pointsPerSeason_randomIntercepts)
deviance(pointsPerSeason_positionAgeFixedLinearSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction)
```

@@

summary(pointsPerSeason_regressionModel)$r.squared
MuMIn::r.squaredGLMM(pointsPerSeason_randomIntercepts)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeFixedLinearSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction)
MuMIn::r.squaredGLMM(pointsPerSeason_gam)

deviance(pointsPerSeason_regressionModel)
deviance(pointsPerSeason_randomIntercepts)
deviance(pointsPerSeason_positionAgeFixedLinearSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction)
deviance(pointsPerSeason_gam)

logLik(pointsPerSeason_regressionModel)
logLik(pointsPerSeason_randomIntercepts)
logLik(pointsPerSeason_positionAgeFixedLinearSlopes)
logLik(pointsPerSeason_positionAgeRandomLinearSlopes)
logLik(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes)
logLik(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction)
logLik(pointsPerSeason_gam)
```

>>>>

### Linear Regression Models {#sec-fantasyPointsByAgeExperienceModelsRegression}

#### Null Model {#sec-fantasyPointsByAgeExperienceModelsRegressionNull}

```{r}
pointsPerSeason_nullModel <- lm(
  fantasy_points ~ 1,
  data = player_stats_seasonal_offense_subset,
  na.action = "na.exclude"
)

summary(pointsPerSeason_nullModel)
summary(pointsPerSeason_nullModel)$r.squared
AIC(pointsPerSeason_nullModel)
MuMIn::AICc(pointsPerSeason_nullModel)
```

#### Linear Model {#sec-fantasyPointsByAgeExperienceModelsRegressionLinear}

```{r}
pointsPerSeason_linearRegression <- lm(
  fantasy_points ~ positionFactor + ageCentered20 + positionFactor:ageCentered20,
  data = player_stats_seasonal_offense_subset,
  na.action = "na.exclude"
)

summary(pointsPerSeason_linearRegression)
summary(pointsPerSeason_linearRegression)$r.squared
AIC(pointsPerSeason_linearRegression)
MuMIn::AICc(pointsPerSeason_linearRegression)
```

#### Quadratic Model {#sec-fantasyPointsByAgeExperienceModelsRegressionLinear}

```{r}
pointsPerSeason_quadraticRegression <- lm(
  fantasy_points ~ positionFactor + ageCentered20 + ageCentered20Quadratic + positionFactor:ageCentered20 + positionFactor:ageCentered20Quadratic,
  data = player_stats_seasonal_offense_subset,
  na.action = "na.exclude"
)

summary(pointsPerSeason_quadraticRegression)
summary(pointsPerSeason_quadraticRegression)$r.squared
AIC(pointsPerSeason_quadraticRegression)
MuMIn::AICc(pointsPerSeason_quadraticRegression)
```

#### Compare Models {#sec-fantasyPointsByAgeExperienceModelsRegressionCompare}

```{r}
anova(
  pointsPerSeason_linearRegression,
  pointsPerSeason_quadraticRegression
)

AIC(
  pointsPerSeason_nullModel,
  pointsPerSeason_linearRegression,
  pointsPerSeason_quadraticRegression
  )

lmModels <- list(
  "nullModel" = pointsPerSeason_nullModel,
  "linearRegression" = pointsPerSeason_linearRegression,
  "quadraticRegression" = pointsPerSeason_quadraticRegression
)

bbmle::AICtab(lmModels)

summary(pointsPerSeason_nullModel)$r.squared
summary(pointsPerSeason_linearRegression)$r.squared
summary(pointsPerSeason_quadraticRegression)$r.squared

deviance(pointsPerSeason_nullModel)
deviance(pointsPerSeason_linearRegression)
deviance(pointsPerSeason_quadraticRegression)

logLik(pointsPerSeason_nullModel)
logLik(pointsPerSeason_linearRegression)
logLik(pointsPerSeason_quadraticRegression)
```

@@

@@

After fitting several models, we now must compare their fit to determine which model fits "best" while also considering parsimony.
Parsimonious models are more likely to be true and more likely to generalize to other samples, because more complex models are more likely to overfit the data.
Thus, more complex models will almost always fit better than simpler models.
Thus, we are not just interested in whether a more complex model fits better than the simpler model; we also care about whether the more complex model fits *significantly* better than the simpler model given its additional complexity.
For evaluating and comparing models, we examine the [likelihood ratio test](#sec-fantasyPointsByAgeExperienceModelsAgeCompareLRT), the [Akaike Information Criterion](#sec-fantasyPointsByAgeExperienceModelsAgeCompareAIC) (AIC), the [corrected AIC](#sec-fantasyPointsByAgeExperienceModelsAgeCompareAICc) (AICc), the [Bayesian Information Criterion](#sec-fantasyPointsByAgeExperienceModelsAgeCompareBIC) (BIC), [$R^2$](#sec-fantasyPointsByAgeExperienceModelsAgeCompareRsquared), [deviance](#sec-fantasyPointsByAgeExperienceModelsAgeCompareDeviance), and [log likelihood](#sec-fantasyPointsByAgeExperienceModelsAgeCompareLogLik).

@@

```{r}
lmVsMixedModel <- list(
  "nullModel" = pointsPerSeason_nullModel,
  "randomIntercepts" = pointsPerSeason_randomIntercepts
)

lmAndMixedModels <- list(
  "nullModel" = pointsPerSeason_nullModel,
  "linearRegression" = pointsPerSeason_linearRegression,
  "quadraticRegression" = pointsPerSeason_quadraticRegression,
  "randomIntercepts" = pointsPerSeason_randomIntercepts,
  "position" = pointsPerSeason_position,
  "fixedLinear" = pointsPerSeason_positionAgeFixedLinearSlopes,
  "randomLinear" = pointsPerSeason_positionAgeRandomLinearSlopes,
  "randomLinearFixedQuadratic" = pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes,
  "randomLinearFixedQuadraticInteraction" = pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction
)

@@

```{r}
AIC(
  pointsPerSeason_nullModel,
  pointsPerSeason_linearRegression,
  pointsPerSeason_quadraticRegression,
  pointsPerSeason_randomIntercepts,
  pointsPerSeason_positionAgeFixedLinearSlopes,
  pointsPerSeason_positionAgeRandomLinearSlopes,
  pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes,
  pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction
  )

@@

```{r}
BIC(
  pointsPerSeason_nullModel,
  pointsPerSeason_linearRegression,
  pointsPerSeason_quadraticRegression,
  pointsPerSeason_randomIntercepts,
  pointsPerSeason_positionAgeFixedLinearSlopes,
  pointsPerSeason_positionAgeRandomLinearSlopes,
  pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes,
  pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction
  )

@@

AICcmodavg::bictab(mixedModels)
#bbmle::AICctab(mixedModels) # throws error (different numbers of observations)

@@

```{r}
summary(pointsPerSeason_nullModel)$r.squared
summary(pointsPerSeason_linearRegression)$r.squared
summary(pointsPerSeason_quadraticRegression)$r.squared
MuMIn::r.squaredGLMM(pointsPerSeason_randomIntercepts)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeFixedLinearSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction)
```

@@

```{r}
deviance(pointsPerSeason_nullModel)
deviance(pointsPerSeason_linearRegression)
deviance(pointsPerSeason_quadraticRegression)
deviance(pointsPerSeason_randomIntercepts)
deviance(pointsPerSeason_positionAgeFixedLinearSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction)
```

###### Log Likelihood {#sec-fantasyPointsByAgeExperienceModelsAgeCompareLogLik}

```{r}
logLik(pointsPerSeason_nullModel)
logLik(pointsPerSeason_linearRegression)
logLik(pointsPerSeason_quadraticRegression)
logLik(pointsPerSeason_randomIntercepts)
logLik(pointsPerSeason_positionAgeFixedLinearSlopes)
logLik(pointsPerSeason_positionAgeRandomLinearSlopes)
logLik(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes)
logLik(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction)
```

@@

summary(pointsPerSeason_nullModel)$r.squared
summary(pointsPerSeason_linearRegression)$r.squared
summary(pointsPerSeason_quadraticRegression)$r.squared
MuMIn::r.squaredGLMM(pointsPerSeason_randomIntercepts)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeFixedLinearSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes)
MuMIn::r.squaredGLMM(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction)
MuMIn::r.squaredGLMM(pointsPerSeason_gam)

deviance(pointsPerSeason_nullModel)
deviance(pointsPerSeason_linearRegression)
deviance(pointsPerSeason_quadraticRegression)
deviance(pointsPerSeason_randomIntercepts)
deviance(pointsPerSeason_positionAgeFixedLinearSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes)
deviance(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction)
deviance(pointsPerSeason_gam)

logLik(pointsPerSeason_nullModel)
logLik(pointsPerSeason_linearRegression)
logLik(pointsPerSeason_quadraticRegression)
logLik(pointsPerSeason_randomIntercepts)
logLik(pointsPerSeason_positionAgeFixedLinearSlopes)
logLik(pointsPerSeason_positionAgeRandomLinearSlopes)
logLik(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopes)
logLik(pointsPerSeason_positionAgeRandomLinearFixedQuadraticSlopesInteraction)
logLik(pointsPerSeason_gam)
```
